Space astrometry of the very massive $\sim$$150\,M_\odot$ candidate runaway star VFTS682

### 2018-09-20 MNRASformat/referee_report_answer.txt

I have read the paper entitled "Space astrometry of the very massive ~ 150 M_sun candidate runaway star VFTS682” by Renzo, Mathieu et al. The paper focusses on the astrometry of the widely-discussed, massive, Wolf-Rayet star VFTS682 in the 30 Dor region. Although this particular very massive star (VMS) has been on focus for several years now since its identification by the VFTS survey and classification as a candidate runaway VMS from the central young massive cluster R136 (Bestenlehner et al. 2011), the present work is a significant leap forward in this candidature as it takes into account, for the first time, the latest proper motion measurements of this star from Gaia and HST. Given that VFTS682 is widely discussed in several pressing contexts such as the production mechanisms of runaway VMSs, the possibility of isolated massive star formation, and the estimation of VMSs’ ages, I think this paper is worth publishing as a letter. Overall, I find the paper well presented and concise. Before the acceptance, however, I recommend the following improvements in the paper.

We thank the referee for their careful reading of our manuscript, and for the suggestions below. The changes in the text are highlighted in green bold font.

1. Sec. 1,  paragraph 4: in the sentence beginning with "N-body simulations indicate that…”, it would be clearer to add the phrase “through dynamical ejections” after the word “expected”.

For the sake of conciseness, we have added "N- body simulations indicate that the *dynamical* ejection of...", which we believe achieves the clarification suggested by the referee with less words.

2. It is not immediately clear to me how the widths of the green vertical bands in Fig. 1 are estimated. It is important to explain this in the text - these bands are important to demonstrate the likeliness of VFTS682 for being an early dynamical runaway from R136. Perhaps similar bands can also be indicated for VFTS16 and VFTS72 in the panels of Fig. 1, for comparing their cases with that of VFTS682.

The green vertical band in the top panel is determined by the error bars on the apparent age of the star: the expected proper motion is the projected distance between the cluster and the star (~29pc) divided by the apparent age of the star (1+/-0.2Myr) which gives the v2D reported in Tab. 2. For the bottom panel we have assumed a width of 45 degrees centered in theta=0. We have added a clarification in the main text of Sec. 3 and a pointer to it in the figure caption.

3. I find Fig. 2 somewhat crowded with lines, especially around VFTS682. For improving the visibility, I suggest that either the central value can be back-traced with dashed lines (say) or the error cones can be shown with dashed lines (preserving the corresponding colours).

We have marked the error cones with dot-dashed lines in the updated version. We have also changed the color of the HST arrows to increase the contrast with the background figure.

4. Fig. 1 is in the top-bottom mode, not left-right. This should be rectified in the first sentence of Sec. 3.

Thank you for spotting this, we have corrected the mistake

5. Sec. 4, paragraph 3: in regard to the statement "N-body dynamics typically ejects the least massive star among those interacting…”, it is perhaps useful to remind the reader that the dynamical ejection fraction, nevertheless, overall increases with increasing stellar mass, as demonstrated in, e.g., Banerjee S. et al., 2012, ApJ, 746, 15 (see their Fig. 4), since the most massive entities would be most centrally segregated and hence interact most efficiently.

We have added a clarifying remark before the citation.

6. Sec. 4, paragraph 4: "In their model, they relied on (dynamically driven) stellar mergers to explain the high masses of VFTS682 and the massive members of R136.” Note that the dynamical merger scenario is not instrumental in explaining members (either residing in or dynamically ejected from R136) up to initial stellar masses of 150 M_sun, due to the adopted IMF in the models of Banerjee et al. (2012) that extends up to 150 M_sun. It is necessary to explain any single star (in-cluster or ejected) beyond 150 M_sun, given these models: see Banerjee S. et al., 2012, MNRAS, 426, 1416 and Crowther P. A. et al., 2010, MNRAS, 408, 731. (In case > 150 M_sun stars would also derive directly from massive-star-formation process, stellar mergers would not be instrumental also for such “super-canonical” stars - for such stars, the dynamical ejection fraction approaches unity as in Banerjee, S. et al., 2012, ApJ, 746, 15 and Fujii, M. S., Portegies Zwart S., 2011, Science, 334, 1380.) Therefore, I suggest rephrasing the quoted sentence accordingly.

We agree with the referee on this point. Our intent in emphasizing the importance of mergers in the simulations from Banerjee et al. is to implicitly pose the question of whether these VMS are mergers or not. While the mergers are not instrumental in explaining their results, the fact that using an IMF up to 150Msun only already gave them enough VMS stars both in the cluster and ejected around it suggest that if they had allowed for the formation of even more massive stars, then the efficiency of dynamically driven mergers together with initially VMS might possibly result in an over-prediction of VMS in the region. We think that this is an interesting point were observations can put constraints on the physics of cluster dynamics and stellar mergers and would prefer to keep the sentence as it is.

7. Sec. 4, second to last paragraph: I recommend also citing the recent work Belczynski, K. et al., 2016, A&A, 594, A97 in the context of the (im)possibility of avoiding pair-instability SN, as invoked there. According to the above work, pair-instability SN can easily be avoided at the LMC metallicity for at least up to 150 M_sun ZAMS (the initially approx. 150 M_sun VFTS682 can easily be a non-merger product; see point 6).

The reference suggested is not a primary one (the work of Belczynski et al. is based on the detailed simulations of Woosley 2017, http://adsabs.harvard.edu/abs/2017ApJ...836..244W), we have added a citation to the primary source instead.

### 2018-09-20 MNRASformat/VFTS682_MNRASL.tex
comments from Eva, Chris, and referee +  report draft

In has been proposed that most, if not all, stars form in clusters
\citep[and references therein]{lada:03}. In this picture, field stars are primarily the result
of the dissolution of dense groups.  However, a significant population
of massive stars exists in relative isolation, far from dense clusters
or OB associations and their origin remains a matter of debate
\citep{gvaramadze:12, lamb:16,ward:18}. One hypothesis to explain
the population of relatively isolated massive stars is that they
formed in the field \citep[e.g.,][]{parker:07}. Another
hypothesis is that these massive stars were ejected from the clusters
in which they formed. Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the companion  star \citep[e.g.,][]{blaauw:61, renzo:18}.

\vspace*{-9pt}
% A contribution to the debate on whether or not massive
% stars form in relative isolation was presented by
% \cite{bestenlehner:11} and \cite{bressert:12}, who discussed the case of the very massive star
% VFTS682.
One of the most extreme examples that has been considered in this
debate is the very massive star VFTS682  \citep[][]{bestenlehner:11,
  bressert:12}. This star is located in the field of the 30 Doradus
(30Dor) region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $137.8^{+27.5}_{-15.9}\,M_\odot$ corresponding to an initial mass of $150.0^{+28.7}_{-17.4}\,M_\odot$
\citep{schneider:18}. 
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
From the spectral point of view, it is reminiscent of the very
massive stars in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
In particular, a remarkable similarity exist between the
spectra of VFTS682 and R136a3 \citep{rubio-diez:17}.%  for which \citet{crowther:16} report a
% current mass estimate of $180^{+30}_{-30}\Msun$.

VFTS682 stands out by its relative isolation at a projected distance of 119.4 arcseconds, corresponding to 
$29$\,pc, from  the star cluster R136. \citet{bestenlehner:11}
considered two possible explanations for the offset: either
the star formed in situ as an isolated massive star, or it was ejected from  R136. N-body simulations 
indicate that the ejection of very massive stars like VFTS682 is
expected \citep[e.g.][]{fujii:11, banerjee:12}. The capability of
young clusters of ejecting a large number of (very) massive stars is supported by
the recent findings of proper motion studies \citep[e.g.,][]{lennon:18, drew:18}.

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} (J2000), was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the most luminous stars in the region. The absence of periodic radial
velocity (RV)
variations suggests that the star is unlikely to have close binary
companions \citep[][]{bestenlehner:11}, unless the orbital inclination is
very high. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the BONNSAI code \citep{schneider:14,schneider:17} provide
estimates for the age, present mass and initial mass, % indicating that
% this star among the most massive stars known,
% see
as given in \Tabref{tab:star_param}. % for an overview of the parameters.

@@

Estimates of the radial velocity are complicated by the variable,
possibly inhomogeneous, optically thick wind  typical of emission line
stars.  We therefore caution
against over-interpreting the radial velocities estimates. \citet{bestenlehner:11} estimate a mass loss rate of
$\log_{10}(\dot{M}/[M_\odot \ \mathrm{yr}^{-1}])=-4.1\pm0.2$, not accounting for the
possible effect of clumping.  % , but we will summarize the estimates that have been published. 
They estimate a RV of
$300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which
is offset from the average radial velocity of the region of
$270\pm10\kms$. This was suggested as indicating a runaway nature, but
it is no proof of it. % , and more recent measurements based on X-shooter
% data give different estimates for the RV ({\color{blue} Rubio-D{\'
%     i}ez et al. in prep}).
\cite{bressert:12} note an offset compared between the RV of
the star and the nebular lines from the gas filaments in its
vicinity. This is consistent with the expectation if the star was not
formed in situ. We will refrain from using the RV measurements in this
work, and focus on the velocities on the plane of the sky.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}\\
  \vspace*{-5pt}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  % \caption{\emph{Top panel}: distribution of OB-type and Wolf-Rayet stars in proper
  %   motion relative to R136 from \emph{Gaia} DR2. VFTS682 is not an outlier, but
  %   its relative proper motion matches the expected value (green
  %   vertical band) if it were indeed
  %   ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. %The top axis shows the conversion to $\kms$
  %   % assuming a distance of 50\,kpc.
  %   \emph{Bottom panel}:  distribution of
  %   angles between the relative proper motion direction and the radial
  %   direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
  %   the best value is in agreement with the hypothesis of dynamical
  %   ejection (green vertical band). The peak at $\theta\simeq\pi/2$ is due to stars
  %   belonging to NGC2060. In both
  %   panels, the dark blue histograms contain 317 
  %   stars with error bars smaller than $0.1\,\mathrm{mas \
  %     yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
  %   lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
  %   yr^{-1}}$. }
  \caption{Distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136 (top panel) and proper motion position angle
    (bottom panel), from \emph{Gaia} DR2. Although VFTS682 is not an outlier, its relative
    proper motion matches the value expected for an early dynamical ejection. In both
    panels, the green vertical bands highlight these expectations, the dark blue histograms contain 317 
    stars with error bars smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
    yr^{-1}}$. The peak at $\theta\simeq\pi/2$ in the bottom panel is due to stars
  belonging to NGC2060. }
  \label{fig:dist}
\end{figure}

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/fig2}  
  % \caption{The thick yellow arrows indicate proper motions relative to
  %   R136 (blue
  %   circle) from \emph{Gaia} DR2, multiplied by 0.4\,Myr. The extensions in the opposite
  %   direction are proportional to the apparent ages of the stars, and
  %   the thin lines illustrate the error cone on the potential
  %   origin. The green thick arrow
  %   indicates the HST proper motion of VFTS682, with the relative
  %   error cone and extensions proportional to its apparent
  %   age. The two blue circles indicate the regions of radii 0.01 and 0.05
  %   degrees around the core of R136.}
  \caption{The thick red arrow shows the proper motion
    relative to R136 for VFTS682 from averaging the \emph{Gaia} DR2 and HST
  astrometry, multiplied by 0.4\,Myr. The extension in the opposite
  direction is proportional to the apparent age of the star, and the
  thin lines illustrate the error cone on the potential origin. The
  yellow arrows show the \emph{Gaia} DR2 results alone, and the orange
  arrow the HST results alone. The two blue circles indicate the regions of radii 0.01 and 0.05
   degrees around the core of R136.}  
  \label{fig:main}
\end{figure}

@@

The 30 Dor region was targeted by a two-epoch photometric campaign
with HST providing observations in the F775W filter in October 2011
and October 2014 (GO-12499; P.I.: D.~J.~Lennon). 
\citet{platais:15, platais:18} analyzed the HST data to determine the
relative proper motions and identify candidate runaway stars. The
brightest stars (V$<$14) are saturated in the data set and have been
excluded from the analysis. The high extinction around VFTS682
makes it redder and fainter \citep[$V=16.08$,
$B-V=0.58$,][]{evans:11}, hence it has reasonably accurate HST astrometry
with the WFC3/UVIS camera. This star did not pass a full set of stringent
conditions to be considered as a candidate runaway \citep[][]{platais:18}.
In retrospect, VFTS682 may have been included in the list of likely
OB runaway stars, and it is identified
with the ID source 330375 in their catalog. 
Therefore, their measurements provide useful
complementary estimate of the proper motion of VFTS682 independent from the \emph{Gaia} data.

@@

% We list the relative proper motion in \Tabref{tab:vfts682}. 
\vspace*{-20pt}
\section{The kinematics of VFTS682}
\label{sec:results}

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682 from \emph{Gaia} DR2. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce close to radial ejections, i.e.~$\theta\simeq0$. The green
vertical bands highlight the expectations for these two quantities if VFTS682 were ejected from R136 early in its
lifetime. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS OB-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a runaway as suggested by
\cite{bestenlehner:11}.

@@

Figure~\ref{fig:main} shows the motion of VFTS682 relative to R136
projected on the
sky. We also show VFTS16 and VFTS72 %for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, the thick green
line illustrates the relative proper motion from HST, and the thick
purple line shows the averaged result. The error cone
on the direction of motion is illustrated by the corresponding
extension in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

@@

Based on our results, we consider likely that VFTS682 is the most massive
runaway known to date, with a two-dimensional
projected velocity with respect to R136 of
$38\pm17\kms$ (averaging the \emph{Gaia} DR2 and HST
results). Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.

@@

If the star were ejected dynamically, its isolation makes it an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$ in the inner
cluster, while avoiding
crowding issues. Moreover, 
 its exceptionally large mass raises the question of which stars must populate
the core of the cluster. N-body dynamics typically ejects the least
massive star among those interacting \cite[e.g.,][]{banerjee:12}. Just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136, as it is observed.% which is consistent with the detection
% of extremely massive stars in the core of the
% cluster.

@@

Also \cite{fujii:11} suggest that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other
stars. The spectral similarities between VFTS682 and stars in the core
of R136 are in agreement with this ``bully binary'' model. Interpreting the kinematics of VFTS682 through the lens of their simulations
suggests the presence of a close binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. The kinematic age of VFTS682 puts an
upper limit to the timescale to form the ``bully binary'' in
R136. If identified, such a binary % could be R145 according to \cite{fujii:11}, and it
might be an ideal observational candidate for a dynamically formed progenitor system of
a binary black-hole, provided that stars this massive can avoid a
pair-instability supernova \cite[e.g.,][]{rakavy:67} at LMC
metallicity \citep[see also][]{langer:07}.
Similarly, the final fate of VFTS682 could be either a
pair-instability supernova without compact remnant formation, or
possibly direct collapse to a black hole above the pair-instability
mass gap. The amount of mass loss of these stars will determine their final core
mass and thus their final fate.

VFTS682 is potentially the most massive runaway known to date, and its ejection
from the cluster R136 likely implies that it is only the ``tip of the
iceberg'' of possibly extremely massive runaways in the
region. Studies of this population, enabled by recent and future HST and \emph{Gaia} observations will put constraints on the evolution
of these extreme stars, together with the formation and evolution of
the central cluster itself.
\vspace*{-20pt}
\bibliographystyle{mnras}
\bibliography{bibliography/vfts682}

>>>>

It has been proposed that most, if not all, stars form in clusters
\citep[and references therein]{lada:03}. In this picture, field stars are primarily the result
of the dissolution of dense groups.  However, a \newtext{small but} significant population
of massive stars exists in relative isolation, far from dense clusters
or OB associations and their origin remains a matter of debate
\citep{gvaramadze:12, lamb:16,ward:18}. One hypothesis to explain
the population of relatively isolated massive stars is that they
formed in the field \citep[e.g.,][]{parker:07}. Another
hypothesis is that these massive stars were ejected from the clusters
in which they formed. Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the companion  star \citep[e.g.,][]{blaauw:61, renzo:18}.

\vspace*{-9pt}
% A contribution to the debate on whether or not massive
% stars form in relative isolation was presented by
% \cite{bestenlehner:11} and \cite{bressert:12}, who discussed the case of the very massive star
% VFTS682.
One of the most extreme examples that has been considered in this
debate is the very massive star VFTS682  \citep[][]{bestenlehner:11,
  bressert:12}. This star is located in the field of the 30 Doradus
(30Dor) region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey \citep[VFTS,][]{evans:11}. It is a hydrogen-rich Wolf-Rayet star of spectral type WNh5. Spectral analysis and comparison with evolutionary models lead to an inferred present-day mass of $137.8^{+27.5}_{-15.9}\,M_\odot$ corresponding to an initial mass of $150.0^{+28.7}_{-17.4}\,M_\odot$
\citep{schneider:18}. 
This makes VFTS682 one of the most massive stars known and one of the most extreme objects in the region.
From the spectral point of view, it is reminiscent of the very
massive stars in the core of the R136 cluster \citep{dekoter:97,crowther:10, crowther:16}. 
In particular, a remarkable similarity exists between the
spectra of VFTS682 and R136a3 \citep{rubio-diez:17}.%  for which \citet{crowther:16} report a
% current mass estimate of $180^{+30}_{-30}\Msun$.

VFTS682 stands out by its relative isolation at a projected distance of 119.4 arcseconds, corresponding to 
$29$\,pc, from  the star cluster R136. \citet{bestenlehner:11}
considered two possible explanations for the offset: either
the star formed in situ as an isolated massive star, or it was ejected from  R136. N-body simulations 
indicate that the \newtext{dynamical} ejection of very massive stars like VFTS682 is
expected \citep[e.g.][]{fujii:11, banerjee:12}. The capability of a
young cluster to eject a large number of (very) massive stars is supported by
the recent findings of proper motion studies \citep[e.g.,][]{lennon:18, drew:18}.

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} (J2000), was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measured an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the most luminous stars in the region. The absence of periodic radial
velocity (RV)
variations suggests that the star is unlikely to have close binary
companions \citep[][]{bestenlehner:11}, unless the orbital inclination is
very high. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the BONNSAI code \citep{schneider:14,schneider:17} provide
estimates for the age, present mass and initial mass, % indicating that
% this star among the most massive stars known,
% see
as given in \Tabref{tab:star_param}. % for an overview of the parameters.

@@

Estimates of the radial velocity are complicated by the variable,
possibly inhomogeneous, optically thick wind  typical of emission line
stars.  We therefore caution
against over-interpreting the \newtext{existing} radial velocities estimates. \citet{bestenlehner:11} estimate a mass loss rate of
$\log_{10}(\dot{M}/[M_\odot \ \mathrm{yr}^{-1}])=-4.1\pm0.2$, not accounting for the
possible effect of clumping.  % , but we will summarize the estimates that have been published. 
They estimate a RV of
$300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which
is offset from the average radial velocity of the region of
$270\pm10\kms$. This was suggested as indicating a runaway nature, but
it is no proof of it. % , and more recent measurements based on X-shooter
% data give different estimates for the RV ({\color{blue} Rubio-D{\'
%     i}ez et al. in prep}).
\cite{bressert:12} note an offset compared between the RV of
the star and the nebular lines from the gas filaments in its
vicinity. This is consistent with the expectation if the star was not
formed in situ. \newtext{Given these issues, we refrain} from using the RV measurements in this
work, and focus on the velocities on the plane of the sky.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}\\
  \vspace*{-5pt}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  % \caption{\emph{Top panel}: distribution of OB-type and Wolf-Rayet stars in proper
  %   motion relative to R136 from \emph{Gaia} DR2. VFTS682 is not an outlier, but
  %   its relative proper motion matches the expected value (green
  %   vertical band) if it were indeed
  %   ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. %The top axis shows the conversion to $\kms$
  %   % assuming a distance of 50\,kpc.
  %   \emph{Bottom panel}:  distribution of
  %   angles between the relative proper motion direction and the radial
  %   direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
  %   the best value is in agreement with the hypothesis of dynamical
  %   ejection (green vertical band). The peak at $\theta\simeq\pi/2$ is due to stars
  %   belonging to NGC2060. In both
  %   panels, the dark blue histograms contain 317 
  %   stars with error bars smaller than $0.1\,\mathrm{mas \
  %     yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
  %   lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
  %   yr^{-1}}$. }
  \caption{Distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136 (top panel) and proper motion position angle
    (bottom panel), from \emph{Gaia} DR2. Although VFTS682 is not an outlier, its relative
    proper motion matches the value expected for an early dynamical
    ejection \newtext{(see \Secref{sec:results})}. In both
    panels, the dark blue histograms contain 317 
    stars with error bars smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc and the
    lighter blue histograms contain 36 stars with error bars smaller than $0.05\,\mathrm{mas \
    yr^{-1}}$. The peak at $\theta\simeq\pi/2$ in the bottom panel is due to stars
  belonging to NGC2060. }
  \label{fig:dist}
\end{figure}

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/fig2}  
  \caption{The thick red arrow shows the proper motion
    relative to R136 for VFTS682 from averaging the \emph{Gaia} DR2 and HST
  astrometry, multiplied by 0.4\,Myr. The extension in the opposite
  direction is proportional to the apparent age of the star, and the
  thin lines illustrate the error cone on the potential origin. The
  yellow arrows show the \emph{Gaia} DR2 results alone, and the orange
  arrow the HST results alone. The two blue circles indicate the regions of radii 0.01 and 0.05
   degrees around the core of R136.}  
  \label{fig:main}
\end{figure}

@@

The 30 Dor region was targeted by a two-epoch photometric campaign
with HST providing observations in the F775W filter in October 2011
and October 2014 (GO-12499; P.I.: D.~J.~Lennon). 
\citet{platais:15, platais:18} analyzed the HST data to determine the
relative proper motions and identify candidate runaway stars. The
brightest stars (V$<$14) are saturated in the data set and have been
excluded from the analysis. The high extinction around VFTS682
makes it redder and fainter \citep[$V=16.08$,
$B-V=0.58$,][]{evans:11}, hence it has reasonably accurate HST astrometry
with the WFC3/UVIS camera. This star did not pass a full set of stringent
conditions to be considered as a candidate runaway \citep[][]{platais:18}.
In retrospect, VFTS682 may have been included in the list of likely
OB runaway stars, and it is identified
with the ID source 330375 in their catalog. 
Therefore, their measurements provide \newtext{a} useful
complementary estimate of the proper motion of VFTS682 \newtext{which is} independent from the \emph{Gaia} data.

@@

% We list the relative proper motion in \Tabref{tab:vfts682}. 
\vspace*{-30pt}
\section{The kinematics of VFTS682}
\label{sec:results}

The black points in~\Figref{fig:dist} show the proper motion 
(\newtext{top} panel) and the projected flight direction (\newtext{bottom} panel) relative
to R136 of
VFTS682 from \emph{Gaia} DR2. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce close to radial ejections, i.e.~$\theta\simeq0$. \newtext{The green
vertical bands highlight the expectations for these two quantities. %  if VFTS682 were ejected from R136 early in its
% lifetime.
The width in the top panel is determined
by the error bars on the star's apparent age, and we assume a width of
$45$ degrees in the bottom panel.} For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS OB-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a runaway as suggested by
\cite{bestenlehner:11}.

@@

Figure~\ref{fig:main} shows the motion of VFTS682 relative to R136
projected on the
sky. We also show VFTS16 and VFTS72 %for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, the thick green
line illustrates the relative proper motion from HST, and the thick
purple line shows the averaged result. The error cone
on the direction of motion is illustrated by the corresponding
extension in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
a large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

@@

Based on our results, we consider that VFTS682 is
\newtext{potentially} the most massive
runaway known to date, with a two-dimensional
projected velocity with respect to R136 of
$38\pm17\kms$ (averaging the \emph{Gaia} DR2 and HST
results). Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.

@@

If the star were ejected dynamically, its isolation makes it an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$ in the inner
cluster, while avoiding
crowding issues. Moreover, 
 its exceptionally large mass raises the question of which stars must populate
the core of the cluster. N-body dynamics typically ejects the least
massive star among those interacting \cite[\newtext{although the
  dynamical ejection fraction increases with mass because of mass
  segregation}, e.g.,][]{banerjee:12}. Just
based on the kinematic properties of VFTS682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136, as it is observed.% which is consistent with the detection
% of extremely massive stars in the core of the
% cluster.

@@

Also \cite{fujii:11} suggest that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other
stars. The spectral similarities between VFTS682 and stars in the core
of R136 are in agreement with this ``bully binary'' model. Interpreting the kinematics of VFTS682 through the lens of their simulations
suggests the presence of a close binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. The kinematic age of VFTS682 puts an
upper limit to the timescale to form the ``bully binary'' in
R136. If identified, such a binary % could be R145 according to \cite{fujii:11}, and it
might be an ideal observational candidate for a dynamically formed progenitor system of
a binary black-hole, provided that stars this massive can avoid a
pair-instability supernova \cite[e.g.,][]{rakavy:67} at LMC
metallicity \citep[see also][]{langer:07, woosley:17}.
Similarly, the final fate of VFTS682 could be either a
pair-instability supernova without compact remnant formation, or
possibly direct collapse to a black hole. The amount of mass loss of these stars will determine their final core
mass and thus their final fate.

VFTS682 is potentially the most massive runaway known to date, and its ejection
from the cluster R136 likely implies that it is only the ``tip of the
iceberg'' of massive runaways in the
region. Studies of this population, enabled by recent and future HST and \emph{Gaia} observations will put constraints on the evolution
of these extreme stars, together with the formation and evolution of
the central cluster itself.
\vspace*{-40pt}
\bibliographystyle{mnras}
\bibliography{bibliography/vfts682}
